---
title: "Venice, the Living City"
subtitle: "Interactive highlights from the betweenness analysis of Venice's pedestrian network"
author: "Uri Hershberg & JJ Merelo"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 9)

library(igraph)
library(dplyr)
library(stringr)
library(scales)
library(plotly)
library(DT)
library(visNetwork)

# Categorical palette (slot 1 / slot 8): identity, not magnitude
pal_calle  <- "#2a78d6"
pal_bridge <- "#e34948"
# Sequential blue ramp for magnitude (edge betweenness)
pal_seq <- c("#cde2fb", "#9ec5f4", "#6da7ec", "#3987e5", "#256abf", "#184f95", "#0d366b")

# plotly's histogram and violin traces don't reliably render on a *log-scaled
# axis* (known plotly.js limitation - binning/KDE happens in linear space).
# Workaround: log10-transform the data ourselves and fake log tick labels on
# an otherwise linear axis. Scatter traces (rank plots below) don't need this
# - they support a real log axis natively.
log_tick_vals <- log10(c(0.1, 1, 10, 100, 1000, 10000, 100000, 1000000))
log_tick_text <- c("0.1", "1", "10", "100", "1k", "10k", "100k", "1M")
```

```{r load-data}
load("data/venice.graph.undirected.Rdata")
g <- venice.graph.undirected
E(g)$edge.betweenness <- edge_betweenness(g)

number.of.edges <- ecount(g)
percent.bridges <- round(length(E(g)[E(g)$bridge == TRUE]) * 100 / number.of.edges, 2)
percent.sotoportegos <- round(length(E(g)[E(g)$sotoportego != FALSE]) * 100 / number.of.edges, 2)
```

## TL;DR

This is a condensed, interactive companion to [*"Venice, the Living City"* (WIVACE 2025, Hershberg & Merelo)](https://link.springer.com/chapter/10.1007/978-3-032-33185-4_1#citeas). The paper argues that Venice's resilience as a walkable medieval city can be read directly from the network structure of its pedestrian paths — and specifically from its **bridges (*ponti*)**, the most changeable, easily added-or-removed elements of the network. All charts below are interactive (hover, zoom, filter) and computed live from the same OpenStreetMap-derived graph used in the paper.

Headline findings:

- Venice's foot network has `r format(vcount(g), big.mark=",")` junctions and `r format(ecount(g), big.mark=",")` undirected path segments, of which `r percent.bridges`% are bridges and `r percent.sotoportegos`% are *sotoportegi* (underpasses).
- Bridges have **far higher edge betweenness** than ordinary streets (*calli*) — a sign they were built precisely where the city needed to close a structural gap.
- The two corridors with the single highest betweenness are not the postcard sights (Rialto, San Marco) but two other stretches: one hugging the Bacino di San Marco near the San Zaccaria vaporetto stop, the other running through Campo Sant'Anzolo to the Ponte dell'Accademia.
- Even after removing dead-end alleys, bridges keep a roughly 9x higher median betweenness than calli — the effect isn't an artifact of blind alleys.

## The dataset

Data comes from OpenStreetMap (using Python library [`osmnx`](https://github.com/gboeing/osmnx), place name *"Venezia-Murano-Burano, Veneto Italy"*, downloaded August 2024), restricted to the `walk` network type. The directed OSM graph (14K+ edges) was merged into a single undirected graph, with edges tagged as bridges, tunnels/*sotoportegi*, and by OSM way name and type.

```{r stats-table}
data.frame(
  Metric = c("Junctions (nodes)", "Path segments (edges)", "Bridges", "Sotoportegi (underpasses)"),
  Value = c(format(vcount(g), big.mark = ","), format(ecount(g), big.mark = ","),
            paste0(percent.bridges, "%"), paste0(percent.sotoportegos, "%"))
) %>%
  datatable(rownames = FALSE, options = list(dom = "t", paging = FALSE),
            caption = "Venice pedestrian graph, at a glance")
```

*Note on the figures below:* the original paper illustrates the highest-betweenness corridor with two hand-annotated OpenStreetMap screenshots. The processed graph used here doesn't carry vertex coordinates (only junction IDs survive the OSM → igraph pipeline), so a geographic re-plot isn't possible from this dataset alone. As a substitute we build an interactive, data-driven view of that same corridor below — a subgraph of the highest-betweenness edges, laid out by force-directed layout rather than geography, but computed directly from the analysis rather than annotated by hand.

## Which edges matter most?

Edge betweenness (Newman & Girvan) measures how many shortest paths between all pairs of junctions pass through a given edge — a proxy for how much a single street or bridge holds the city together.

```{r top-table}
edge.betweenness.table <- data.frame(
  edge = sapply(E(g)$name, function(x) x[1]),
  betweenness = E(g)$edge.betweenness,
  type = ifelse(E(g)$bridge, "Bridge", "Calle"),
  stringsAsFactors = FALSE
)
sorted.table <- edge.betweenness.table[order(-edge.betweenness.table$betweenness), ]
sorted.table$rank <- seq_len(nrow(sorted.table))

datatable(
  head(sorted.table[, c("rank", "edge", "betweenness", "type")], 20),
  colnames = c("Rank", "Edge name", "Betweenness", "Type"),
  rownames = FALSE,
  options = list(pageLength = 10, dom = "ftip"),
  caption = "Top 20 edges ranked by betweenness centrality"
) %>%
  formatRound("betweenness", 1) %>%
  formatStyle("type", target = "row",
              backgroundColor = styleEqual(c("Bridge", "Calle"), c("#fbe4e3", "transparent")))
```

### The highest-betweenness corridor, as a network

```{r subnetwork}
top_n <- 40
top_idx <- order(E(g)$edge.betweenness, decreasing = TRUE)[seq_len(top_n)]
sub_g <- subgraph.edges(g, E(g)[top_idx], delete.vertices = TRUE)

sub_edge_names <- sapply(E(sub_g)$name, function(x) x[1])
edgelist <- as_edgelist(sub_g, names = FALSE)

edges_df <- data.frame(
  from = edgelist[, 1],
  to = edgelist[, 2],
  color = ifelse(E(sub_g)$bridge, pal_bridge, pal_calle),
  width = rescale(E(sub_g)$edge.betweenness, to = c(2, 12)),
  title = paste0("<b>", sub_edge_names, "</b><br>",
                 ifelse(E(sub_g)$bridge, "Bridge", "Calle"),
                 "<br>Betweenness: ", round(E(sub_g)$edge.betweenness))
)
nodes_df <- data.frame(id = seq_len(vcount(sub_g)), label = "", shape = "dot",
                        size = 6, color = "#898781")

visNetwork(nodes_df, edges_df, height = "520px") %>%
  visEdges(smooth = FALSE) %>%
  visOptions(highlightNearest = list(enabled = TRUE, degree = 1)) %>%
  visLayout(randomSeed = 42) %>%
  visLegend(addEdges = data.frame(label = c("Bridge", "Calle"), color = c(pal_bridge, pal_calle)),
            useGroups = FALSE, width = 0.15)
```

Hover an edge for its name and betweenness; click a node to highlight its immediate neighborhood. The two clusters described in the paper are visible here as the two denser knots: one anchored on the *Ponte del Sepolcro – San Zaccaria – rio dell'Arsenale* stretch, the other on *Campo Sant'Anzolo – Campo S. Stefano – Ponte dell'Accademia*.

## Distribution of betweenness

```{r rank-plot}
plot_ly(sorted.table, x = ~rank, y = ~betweenness, type = "scatter", mode = "markers",
        marker = list(color = pal_seq[4], size = 5, opacity = 0.55),
        text = ~edge,
        hovertemplate = "%{text}<br>Rank %{x}<br>Betweenness %{y:.0f}<extra></extra>") %>%
  layout(title = "Rank-value plot of edge betweenness",
         xaxis = list(title = "Rank"),
         yaxis = list(title = "Edge betweenness (log scale)", type = "log"))
```

The broken-stick shape — a slow decline, a long plateau, then a fast drop — is visible directly: hover any point to see which street or bridge it is. The plateau (a few thousand edges around the same value) corresponds mostly to dead-end *rami* and *corti*.

```{r histogram}
sorted.table$log_betweenness <- log10(pmax(sorted.table$betweenness, 0.1))

plot_ly(sorted.table, x = ~log_betweenness, type = "histogram", nbinsx = 40,
        marker = list(color = pal_seq[4], line = list(color = "#fcfcfb", width = 1))) %>%
  layout(title = "Distribution of edge betweenness",
         xaxis = list(title = "Edge betweenness (log scale)",
                      tickvals = log_tick_vals, ticktext = log_tick_text),
         yaxis = list(title = "Count (log scale)", type = "log"))
```

## Bridges vs. *calli*

```{r violin}
bridge.median <- median(edge.betweenness.table$betweenness[edge.betweenness.table$type == "Bridge"])
calle.median  <- median(edge.betweenness.table$betweenness[edge.betweenness.table$type == "Calle"])

edge.betweenness.table$log_betweenness <- log10(pmax(edge.betweenness.table$betweenness, 0.1))

plot_ly(edge.betweenness.table, x = ~type, y = ~log_betweenness, type = "violin",
        color = ~type, colors = c(Bridge = pal_bridge, Calle = pal_calle),
        box = list(visible = TRUE), meanline = list(visible = TRUE), points = FALSE) %>%
  layout(title = "Edge betweenness: bridges vs. calli",
         xaxis = list(title = ""),
         yaxis = list(title = "Edge betweenness (log scale)",
                      tickvals = log_tick_vals, ticktext = log_tick_text),
         showlegend = FALSE)
```

Median edge betweenness is `r round(bridge.median)` for bridges vs. `r round(calle.median)` for calli — bridges run about **`r round(bridge.median / calle.median, 1)`x higher**, evidence that they were placed precisely where the network needed a shortcut.

## Does it survive removing dead ends?

Roughly a third of edges terminate in a degree-1 dead end (blind alleys, closed squares). Dropping those and recomputing betweenness on the pruned graph:

```{r prune}
blind.alleys <- which(degree(g) <= 1)
g.pruned <- delete_vertices(g, blind.alleys)
E(g.pruned)$edge.betweenness <- edge_betweenness(g.pruned)

pruned.df <- data.frame(
  betweenness = E(g.pruned)$edge.betweenness,
  type = ifelse(E(g.pruned)$bridge, "Bridge", "Calle")
) %>%
  group_by(type) %>%
  arrange(desc(betweenness)) %>%
  mutate(rank = row_number()) %>%
  ungroup()

plot_ly(pruned.df, x = ~rank, y = ~betweenness, color = ~type,
        colors = c(Bridge = pal_bridge, Calle = pal_calle),
        type = "scatter", mode = "markers", marker = list(size = 5, opacity = 0.55)) %>%
  layout(title = "Rank-value plot after removing dead ends, by type",
         xaxis = list(title = "Rank (within type)"),
         yaxis = list(title = "Edge betweenness (log scale)", type = "log"))
```

```{r prune-medians}
bridge.median.pruned <- median(pruned.df$betweenness[pruned.df$type == "Bridge"])
calle.median.pruned  <- median(pruned.df$betweenness[pruned.df$type == "Calle"])
```

After pruning, median betweenness is `r round(bridge.median.pruned)` for bridges vs. `r round(calle.median.pruned)` for calli — still about **`r round(bridge.median.pruned / calle.median.pruned, 1)`x higher**. The plateau seen above largely disappears (most plateau edges were the dead ends just removed), but the bridge advantage holds.

## Street types and their connective power

Not all non-bridge ways are alike either. Grouping edges by their first OSM street-name word (*calle*, *fondamenta*, *campo*, *rio*, *salizada*, ...):

```{r type-table}
calle.type <- data.frame(
  name = sapply(E(g)$name, function(x) x[1]),
  betweenness = E(g)$edge.betweenness,
  stringsAsFactors = FALSE
)
calle.type$edge.type <- str_extract(calle.type$name, "\\w*")

type.summary <- calle.type %>%
  group_by(edge.type) %>%
  filter(n() > 50, edge.type != "") %>%
  summarize(edges = n(), median_betweenness = median(betweenness)) %>%
  arrange(desc(median_betweenness))

datatable(type.summary, colnames = c("Segment name", "Occurrences", "Median betweenness"),
          rownames = FALSE, options = list(pageLength = 10, dom = "ftip"),
          caption = "Street/way types (>50 occurrences), ranked by median edge betweenness") %>%
  formatRound("median_betweenness", 1)
```

Four of the top five types by median betweenness — *riva*, *ponte*, *rio* (paved-over canal), *fondamenta* — are tangent to or built directly over water. The fifth, *salizada* ("paved"), marks the streets deliberately paved over open ground as early as the 13th century. Both are signatures of the same process: edges added, over centuries, exactly where they closed a structural gap in the network.

## Takeaways

- Venice's most connective paths are disproportionately its most **changeable** ones — bridges, paved-over canals, waterside platforms — not its oldest or most iconic streets.
- The two single highest-betweenness corridors sit near, but not on top of, San Marco and the Rialto — likely reflecting later adaptations (the Accademia bridge and its academic district; the residential expansion of Castello around the Biennale gardens) rather than the historical mercantile core.
- The pattern survives pruning dead ends, so it isn't an artifact of Venice's many blind alleys.
- Read together, this is consistent with Venice behaving as a *living* system: not statically stable, but stable *because* its network keeps adapting.

For the full argument, related work, and discussion, see the WIVACE 2025 paper this notebook summarizes.

---

<small>Source data: `data/venice.graph.undirected.Rdata` (OpenStreetMap, August 2024 extract). Rendered with `igraph`, `plotly`, `DT`, and `visNetwork` — all HTML widgets, no static/PDF plots.</small>
